Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-06-11 PhDThesis.tex
changes after proof reading

Rapidly evolving pathogens pose an ever increasing threat to the human population.
This threats are exemplified by recent infections with H5N1 avian influenza A, pandemic spread of swine-origin H1N1 influenza \citep{Rambaut2008} or Severe Acute Respiratory Syndrome (SARS) \citep{Ksiazek2003}.
Unlike zoonotic diseases human RNA viruses have no natural barriers that could hamper their escalation.
This is due to the extensive nature of human mobility \citep{Brockmann2006}.

Despite the ever increasing impact of emerging infectious diseases (EIDs) the first studies combining molecular evolution with geographical dispersal are less than 50 years old \citep{Henning1966}. 
Most of these pioneering studies were limited to mapping phylogenetic tip taxa to their geographical coordinates \citep{Kidd2006} and comprehensive quantitative frameworks have only recently begun being developed.

Models that combined the process of spatial diffusion with genetic substitution are only as recent as this decade \citep{Lemey2009,Lemey2010}.
These advances are mostly hampered by the computational burden that they impose.
Finding the maximum likelihood tree requires a search through the space of all possible tree topologies, which quickly becomes computationally intractable, as depicted in Table \ref{tab:maxLikeBurden}.

\begin{table}
\caption{{\bf{Number of rooted trees per number of taxa.}} $4 \times 10^{80} $ is the no of atoms in the observable universe.}
\begin{centering}
\begin{tabular}{cc}
% {@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
Tips & Rooted trees $\frac{(2N-3)!}{2^{N-2}\cdot(N-2)!}$ \tabularnewline
\hline 
\rowcolor{gray1}
$3$ & $3$\tabularnewline
$6$ & $945$\tabularnewline
\rowcolor{gray1}
$9$ & $2'027'025$\tabularnewline
$15$ & $2\times10^{14}$\tabularnewline
\rowcolor{gray1}
$20$ & $8\times10^{21}$\tabularnewline
$55$ & $3.19\times10^{86}$\tabularnewline
\end{tabular}
\par\end{centering}
\label{tab:maxLikeBurden}
\end{table}

Adopting Bayesian statistical approach, implemented in a flexible framework of molecular evolution \citep{Drummond2012} helps circumvent some of these hindrances by focusing sampling efforts on the topologies 
% parts of the tree space
that best fit the data.
Nonetheless Bayesian phylogenetic analyses remain highly computationally intensive and a vast area of research is now aiming at bringing new computational advances to actively stretch these limits \citep{Suchard2009}.

@@

Central assumption of viral phylogenetics is that thee pidemiological processes that shape the fitness of viral populations leave a measurable footprint on their genomes.
Due to the rapid nature of their evolution this process of character substitution can be observed in almost real time, and gives the opportunity to sample genetic data at different points in time, as opposed to ultrametricly sampled data (\ref{fig:ultrametric}).

%---ULTRAMETRIC TREE---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.35]{ultrametric} 
\end{center}
\caption{{ \footnotesize {\bf Different sampling schemes.} 
Tree on the right has all of it's sampling dates set at $t=0$. Conversly the tree on the left is sampled at different points in time.
}}
\label{fig:ultrametric}
\end{figure}

Such data-sets are usually reffered to as time-heterogeneous and frequently arise when from studies involving viruses which accumulate statistically significant amounts of genetic change in a relatively short amount of time.
Populations for which such data can be obtained have been called measurably evolving populations (MEPs, cite{Drummond2003}) and have been instrumental in providing  estimates of the evolutionary rate and in advances to the molecular clock models \citep{Drummond2006}.

In studies known as phylodynamics the proceses which shape th eviral genome are integrated with epidemiological dynamics, by assuming that both the evolution and 
geographical dispersal are independent, stochastic processes which unfold on phylogenetic trees.
In that sense the tree is not only a record of evolutionary ancestry, but also of dispersal in actual geographical coordinates, which happen simultaneously.
Our research group has conducts research which aims at developing a full quantitative understanding of the processes that shape the epidemiology and evolution of MEP-like viruses.

% TODO: this is a silly section name; alter
\section{Model extension, simulation and visualisation}

% models
Generally phylogenetic analysis employing stochastic processes make strong restrictive assumptions regarding the underlying substitution process.
They are mainly posited to reduce the number of free parameters of the models describing the molecular evolution and ease both the already mentioned computational tractability and mathematical tractability.
Unfortunately many of these assumptions hamper the application of the methods to real-world, complex data-sets.
Relaxing some of this assumptions and development of more biologically plausibe models, able to grasp the complex nature of phylodynamic dispersal processes is one the aims of the presented thesis \citep{Bielejec2014a}.
Another goal is to make the general applicability suitable for phylogeographic problems as well as beyond them, with potential to turn the findings into usefull predictions of the emergence of infectious diseases, e.g. by identifying the sources and sinks of the geographical dispersal.

% simuation
With the development of novel phylogenetic inference methods comes the need to synthesize evolutionary data in order to compare estimator performance and to characterize strengths and weaknesses of different approaches (e.g. \cite{Arenas2012}, \cite{Hoban2011}).
Whereas the true underlying evolutionary relationships between biological sequences are generally unknown, Monte Carlo simulations allow generating test scenarios while controlling for the evolutionary history as well as the tempo and mode of evolution. 
Part of this thesis is the development and introduction of a flexible utility software for for sequence simulation, that integrates both tree-generative processes as well as models responsible for shaping the molecular histories and geographical dispersal \citep{bielejec2014}.

% visualisation
Phylogenetic inference unevitably results in complex, multi-dimensional output data.
This dimensionality is further incremented by adding the geography to the usuall time-component.
Finally, Bayesian analysis result in posterior distributions of trees, with each tree containing estimated node locations and other traits of interest.
These aspects represent a major challenge for visualizing the inference. 
Yet clear and intuitive visualizations are an important aspect of every analysis (\cite{Hadley10}).
Spatial phylogenetic projections have been undertaken in the field \citep{Kidd06,Parks09}, yet most of these applications remain limited to mapping phylogenetic tip taxa to their geographical coordinates, without providing robust statistical estimates of the geographic locations at the ancestral nodes nor the measures of uncertainty in the inference.
Fortunately the advent of powerful and flexible Geographic Information Systems (GIS) has fostered an interesting possibility in icorporatin both spatial and temporal information into viral phylogenetic analysis.
Presenting readily interpretable visual summaries of the inferences, that can in turn be communicated to the researchers across different fields is one of the more important aspects of this thesis.
This has been materialized in the development of an user-friendly application for producing compelling and informative statistical visualizations of viral spread in time and space \citep{Bielejec2011}.

@@

Molecular phylogenetic modelling typically assumes that the observed molecular sequence data has been generated by some stochastic process, which emits characters over time.
We can think of stochastic processes as functions of time, which is regarded as a deterministic argument, whose values are random (non-deterministic) variables. 
It is thus a statistical model of a random development, or evolution, in time.
Phylogenetic inference often resorts to processes where time is considered to be continuous and the indexed set of outcomes, called the state space, is countable and finite (discrete).
Thus the stochastic process $X$ is a collection $\left\{ X(t)\; t\in T\right\} $, where each drawn value $X(t)$ is a random variable drawn from state space $\mathcal{E}$, with $K$ possible values.

@@

\begin{equation}
\left\{ P(t)\right\} _{ij}=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}\frac{(r t)^{k}}{k!}e^{-r t}=\underset{k=0}{\overset{\infty}{\sum}}\frac{(\mathbf{R}r t)^{k}}{k!}e^{-r t},
\end{equation}

\noindent
where $\left\{ P(t)\right\} _{ij}$ denotes the $ij\text{-th}$ element of transition probabilities matrix $\mathbf{P}$ over time $t \geq 0$.
Let us denote the matrix $\mathbf{R}-\mathbf{I}$, where $\mathbf{I}$ is a $K \times K$ identity matrix by $\mathbf{Q}$.
Using the power series definition of the transition probability matrix $exp(\mathbf{R})=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}/k!$ we have:

\begin{equation}
\left\{ P(t)\right\} _{ij}=e^{\mathbf{R}r t}e^{-r t} =e^{(\mathbf{R}-\mathbf{I})r t}=e^{\mathbf{Q}r t} 
\label{eq:matrixExp}
\end{equation}

@@

In order to arrive at the finite-time transition probabilities matrix $\mathbf{P}(t)$ for some time $t\geq0$ one needs to compute the matrix exponential in Equation \ref{eq:matrixExp}.
\cite{Moler78nineteendubious} provide an excellent review of numerical methods to compute the matrix exponent, here we will highlight the method utilized by most phylogenetic software, including BEAGLE \citep{ayres2012beagle} and BEAST \citep{Drummond2012}, that uses the Singular Value Decomposition (SVD) of the rate matrix $\mathbf{Q}$.

@@

\noindent where $\mathbf{V}$ is a matrix composed of eigenvectors of $\mathbf{Q}$. 
Multiplying both sides of equation (\ref{eq:diagonalization}) with the edge length $t$ and scaled by the appropriate rate scalar $r$, we can write $\mathbf{Q}$ in form:

@@

\noindent 
where the last transition is due to the fact that exponential of diagonal matrix is obtained by exponentiating every entry on the main diagonal.
We should note here that the tepeated evaluations of $\exp\left(rt\mathbf{Q}\right)$ for different times $t$ are all based on the same SVD of the matrix $\mathbf{Q}$. 
One needs only to scale and exponentiate the eigenvalues to calculate the diagonal matrix $D$ and perform two matrix multiplications, thus reducing all calculations to a simple and effective algorithm.

@@

The elements of $\mathbf{P}$ quantify the finite-time transition probabilities between the $K$ discrete state-space elements.
We have described here a Continuous-time Markov chain (CTMC), that we will use to describe the substitution process at a single site of an alignment.
Every CTMC is completely characterized by its rate matrix $\mathbf{Q}$ and the stochastic matrix $\mathbf{P}(t),\ t\geq0$ is obtained via matrix exponentiation as decibed in Subsection \ref{seq:exponentiation}.

@@

Figure \ref{fig:substitution} depicts an example of such a CTMC substitution model, describing the process of substitution between four discrete nucleotide base pairs, where all the pairs communicate (all states have a non-zero transitional probability of exiting from, i.e. all states are non-absorbing).
Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by:

@@

Unfortunately this assumption clearly does not hold for many real-world data-sets, as the temporal dynamics clearly do shift in time, often in a seasonal, observable manner \citep{Bahl2011}.
One of the goals of the research presented in this thesis is the development of a model that can grasp such situations.

@@

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in states $i\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state of the process.
In other words if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution.

@@

\noindent
Condition \ref{eq:time_reversibility1} is also referred to as the \emph{detailed balance} equation, and posits that the probability of sampling $i$ from the stationary distribution and then going to $j$ over some finite-time $t$ is the same as sampling $j$ and going to $i$. 
% By the definition of $\mathbf{\Pi}$ and $\mathbf{P}$ from (\ref{eq:time_reversibility1}) we have:
% 
% \begin{equation}
% \pi_{i}q_{ij}=\pi_{j}q_{ji}\mathbf{Q}.
% \label{eq:time_reversibility2}
% \end{equation}
% 
% Definition (\ref{eq:time_reversibility1}) and (\ref{eq:time_reversibility2}) are equivalent.
This condition is assumed for strictly computational reasons, i.e. time-reversibility makes it easier to diagonalize $\mathbf{Q}$, which as we have described in 
Subsection \ref{seq:exponentiation} is a strategy frequently employed in phylogenetic software to compute the exponentials of rate matrices.
Time reversibility also simplifies the computation of the likelihood on a tree, since it is then indifferent to the position of the root, which we will discuss in Subsection \ref{sub:likelihood}.

@@

In this subsection we will discuss the simplest model of nucleotide substitution, i.e. the JC69 model \citep{Jukes1969}.
The JC69 model assumes that the instantaneous rates of change between every two nucleotides are the same.
We also require that the rat ematrix $Q$ characterizing the model is a stochastic matrix (each row sums to 0).
The infinitestimal rate matrix is thus given by:

@@

\noindent
The finite-time transition probabilities matrix $\mathbf{P}(t)$ for time $t>0$ is calculated using matrix exponentiation  (see subsection \ref{sub:exponentiation}):

@@

%TODO: confounded time & rate
\noindent
Second, the rate and time in (\ref{eq:jc69Finite}) occur as a product, which means that they are confounded.
If we were to simulate two sequences, using Algorithm listed in (\ref{alg:simulation}) and starting the random number generators form the same seed, they will look the same for every $r$ and $t$ as long as their product $rt$ is the same.
This means that typically any inference performed is focused on inferring the product parameter $\theta=rt$.

@@

To gain a better understanding of the process shaping the data observed at the tips of a phylogenetic tree let us consider the probelm of simulating a sequence evolution on a tree.
We start with a first character $i$ sampled at the root of the tree $x_0$ from the stationary distribution $\mathbf{\Pi}$.
Let us assume that two directly neighboring nodes are $x_j$ and $x_k$.
The character sampled at the root waits an Exponentialy distributed amount of time $t\geq0$ (see Figure \ref{fig:poisson}) before moving down the tree to one of the children nodes $x_i$.
Such traversal is said to be following the pre-order, i.e. parental nodes are visited before child nodes.   
The process then randomly decides by a coin flip to which of the two neighbouring nodes to move next.
Let us assume that the process moved to $x_j$.
The state $j$ for the node $x_j$ is sampled conditional on the value $i$ already sampled at the root using the branch length expressed in real time $t$ and the vector probabilities in the row $i$ of the finite-time probabilities matrix $\mathbf{P}(t)$, calculated using SVD of the rate matrix $\mathbf{Q}$ that characterizes the CTMC.

@@

This constitutes the Markov property of a CTMC, as defined in (\ref{eq:markov}).
The sampling is then repeated on the node $x_j$, before recursively proceeding to the other trio of nodes.
The states sampled at the tips of the tree constitute the character sequence for one site of the alignment.
In thi slight the CTMC process can be viewed as a continuous-time random-walk, unfolding on the topology $\mathbf{F}$, with it's behaviour described by a rate-matrix $\mathbf{Q}$.

@@

Likelihood is defined as a conditional probability of observing the data given the parameters, in a function of those parameters.
Discrete molecular sequence data comes in a form of aligned matrix $\mathbf{X}$ of size $n \times l$, where $x_{jh}$ is the $h$-th character in $j$-th sequence.
Single column $\mathbf{x}_{h}$ of data matrix constitutes a single site.
Like before the standard assumption of independence posits that different sites evolve independently and that once two branches split at a node the evolution on those is also independent.

In Subsection \ref{sub:evolutionOnTree} we described how some process $\mathbf{Q}$ gives rise to the observed data $\mathbf{X}$.
In the typicall situation we start with just the observed data and aim the inference efforts at modelling the process that generated the alignment.
We can therefore denote the likelihood simply as:

@@

In this thesis we will not discuss the maximum likelihood estimation itself, rather focus on closely related methods of Bayesian inference.
The likelihood is nontheless required for both methods of inference, hence we will now discuss how to calculate the likelihood of observing a certain realised sequence of characters on a tree, given a substitution model.

@@

To calculate the likelihood means thus to reverse-engineer the evolution as it happened.
Figure \ref{fig:LIKELIHOOD} presents single site $TCACC$ observed at the tips of the tree.
External nodes are numbered $x_{1}, \ldots x_{5}$, internal nodes $x_{6}, \ldots x_{8}$ and the root is placed at the node $x_{0}$.
Length of each branches leading to node $i$ is denoted as $t_{i}$.
Using methods presented in Subsection \ref{sub:exponentiation} we calculate the finite-time transition probabilities $p_{ij}(t_{i})$ between all pairs of characters in the alignment.
According to the independence assumptions to calculate the probability at a single site we need to sum over all possible combinations that could give rise to the observed sequence. Thus:

\begin{equation}
\underset{X_{0}\in S}{\sum}\underset{X_{6}\in S}{\sum}\underset{X_{7}\in S}{\sum}\underset{X_{8}\in S}{\sum}[\pi_{x_{0}}\cdot p_{x_{0}x_{8}}(t_{7})\cdot p_{x_{0}x_{7}}(t_{8})\cdot p_{x_{8}x_{6}}(t_{6})\cdot p_{x_{8}A}(t_{3})\cdot p_{x_{7}C}(t_{4})\cdot p_{x_{7}C}(t_{5})\cdot p_{x_{6}T}(t_{1})\cdot p_{x_{6}C}(t_{2})]
\label{eq:likelihoodNaive}
\end{equation}

>>>>

Rapidly evolving pathogens pose an ever increasing threat to the human population.
This threats are exemplified by recent infections with H5N1 avian influenza A, pandemic spread of swine-origin H1N1 influenza \citep{Rambaut2008} or Severe Acute Respiratory Syndrome (SARS) \citep{Ksiazek2003}.
Unlike zoonotic diseases human RNA viruses have no natural barriers that could hamper their escalation.
, mainly due to the extensive nature of human mobility \citep{Brockmann2006}.

Despite the ever increasing impact of emerging infectious diseases (EIDs) the first studies combining molecular evolution with geographical dispersal are less than 50 years old \citep{Henning1966}. 
Most of these pioneering studies were limited to mapping phylogenetic tip taxa to their geographical coordinates \citep{Kidd2006} and comprehensive quantitative frameworks have only recently begun being developed.
Models that combine the process of spatial diffusion with genetic substitution are only as recent as this decade \citep{Lemey2009,Lemey2010}.
The hampered advances in the field of computational phylogenetics is mainly attributed to the computational burden that these compute-intensive calculations impose.
As an example, finding the maximum likelihood tree requires a search through the space of all possible tree topologies, which quickly becomes computationally intractable, as depicted in Table \ref{tab:maxLikeBurden}.

\begin{table}
\caption{{\bf{Number of rooted trees per number of taxa.}} $4 \times 10^{80} $ is the no of atoms in the observable universe.}
\begin{centering}
\footnotesize{
\begin{tabular}{cc}
% {@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
Tips & Rooted trees $\frac{(2N-3)!}{2^{N-2}\cdot(N-2)!}$ \tabularnewline
\hline 
\rowcolor{gray1}
$3$ & $3$\tabularnewline
$6$ & $945$\tabularnewline
\rowcolor{gray1}
$9$ & $2'027'025$\tabularnewline
$15$ & $2\times10^{14}$\tabularnewline
\rowcolor{gray1}
$20$ & $8\times10^{21}$\tabularnewline
$55$ & $3.19\times10^{86}$\tabularnewline
\end{tabular}
}
\par\end{centering}
\label{tab:maxLikeBurden}
\end{table}

Adopting Bayesian statistical approach, implemented in a flexible framework of molecular evolution \citep{Drummond2012} helps circumvent some of these hindrances by focusing sampling efforts on the topologies 
% parts of the tree space
that best fit the data.
Nonetheless Bayesian phylogenetic analyses remain highly computationally intensive and a vast area of research is now aiming at bringing new computational advances to actively stretch these limits \citep{Suchard2009}.
This problem is further aggravated as the advent of high-throughput sequencing produces millions of sequences at low cost resulting in the availability of increasingly large data sets.

%TODO: other problems

@@

Central assumption of viral phylogenetics is that the epidemiological processes that shape the fitness of viral populations leave a measurable footprint on their genomes.
Due to the rapid nature of their evolution, the process of character substitution can be observed in almost real time, and gives the opportunity to sample genetic data at different points in time, as opposed to ultrametricly sampled data such as depicted in Figure \ref{fig:ultrametric}.

%---ULTRAMETRIC TREE---%
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.35]{ultrametric} 
\end{center}
\caption{{ \footnotesize {\bf Different sampling schemes.} 
Tree on the right has all of it's sampling dates set at $t=0$. Conversly the tree on the left is sampled at different points in time and provides temporal information.
}}
\label{fig:ultrametric}
\end{figure}

Such data-sets are usually reffered to as time-heterogeneous and frequently arise from studies involving viruses which accumulate statistically significant amounts of genetic change in a relatively short amount of time.
Populations for which such data can be obtained are called measurably evolving populations (MEPs, \cite{Drummond2003}) and have been instrumental in providing  estimates of the evolutionary rate and in making advances to the molecular clock models \citep{Drummond2006}.

In studies known as phylodynamics the processes which shape the viral genome are integrated with epidemiological dynamics, by assuming that both the evolution and 
geographical dispersal are independent, stochastic processes which unfold on phylogenetic trees.
In that sense the treedpes not only provide a record of evolutionary ancestry, but also of the dispersal in actual geographical coordinates, and both these processes happen simultaneously.
Our group has been involved in research which aims at developing a full quantitative understanding of the processes that shape the epidemiology and the evolution of MEP-like viruses.
In the following section I present the problems on which my involvment in this fascinating area of research has been focused.

\section{Model extension, simulation and visualisation}

% models
Phylogenetic analysis employing stochastic processes make strong restrictive assumptions regarding the underlying substitution process.
These assumptions are mainly posited to reduce the number of free parameters of the models describing the molecular evolution and ease the already mentioned computational and mathematical tractability.
Unfortunately many of these assumptions hamper the application of the methods to the complex, real-world data-sets.
Relaxing some of these assumptions and development of more biologically plausibe models, able to grasp the complex nature of phylodynamic dispersal processes is one the aims of the presented thesis \citep{Bielejec2014a}.
Another goal is to make the general applicability suitable for epidemiological problems as well as beyond them, with potential to turn the findings into usefull predictions of the emergence of infectious diseases, e.g. by identifying the sources and sinks of the geographical dispersal.

% simuation
With the development of novel phylogenetic inference methods comes the need to synthesize evolutionary data, in order to compare estimator performance and characterize strengths and weaknesses of different approaches (e.g. \cite{Arenas2012}, \cite{Hoban2011}).
Whereas the true underlying evolutionary relationships between biological sequences are generally unknown, Monte Carlo simulations allow generating test scenarios while controlling for the evolutionary history as well as the tempo and mode of evolution. 
Part of the research presented in this thesis is devoted towards development of a flexible software for sequence simulation, that integrates both the tree-generative processes, as well as models responsible for shaping the molecular histories and geographical dispersal \citep{bielejec2014}.

% visualisation
Phylogenetic inference unevitably results in complex, multi-dimensional output data.
This dimensionality is further incremented by adding the geography to the usuall evolutionary time component.
Finally, Bayesian analysis result in posterior distributions of trees, with each tree branch annotated with estimated node locations and other traits of interest.
These aspects represent a major challenge for visualizing the inference. 
Yet clear and intuitive visualizations are an important aspect of every analysis \citep{Hadley2010}.
Spatial phylogenetic projections have been already been utilized in the field \citep{Kidd2006,Parks2009}, yet most of these applications remain limited to mapping phylogenetic tip taxa to their geographical coordinates, without providing robust statistical estimates of the geographic locations at the ancestral nodes nor the measures of uncertainty in the inference.
Fortunately the advent of powerful and flexible Geographic Information Systems (GIS) has fostered an interesting possibility to incorporate both spatial and temporal information into viral phylogenetic analysis.
Presenting readily interpretable visual summaries of the inferences, that can in turn be communicated to the researchers across different fields is one of the more important aspects of this thesis.
This has been materialized in the development of an user-friendly application that produces compelling and informative statistical visualizations of viral spread in time and space \citep{Bielejec2011}.

@@

Molecular phylogenetic modelling typically assumes that the observed molecular sequence data has been generated by some stochastic process, which emits characters over time.
We can think of stochastic processes as functions of time, which is regarded as a deterministic argument, whose values are random (non-deterministic) variables. 
It is thus a statistical model of a random development, or evolution, in time.
Phylogenetic inference often resorts to processes where time is considered to be continuous and the indexed set of outcomes, called the state space, is countable and finite (discrete).
% The stochastic process $X$ is a collection $\left\{ X(t):\; t\in T\right\} $, where each drawn value $X(t)$ is a random variable drawn from state space $\mathcal{E}$, with $K$ possible values, indexed by orderer time $T$.

\begin{defn} 
The stochastic process $X$ is a collection $\left\{ X(t):\; t\in T\right\} $, where each drawn value $X(t)$ is a random variable drawn from state space $\mathcal{E}$, with $K$ possible values, indexed by orderer time $T$.
\label{def:stochasticProc}
\end{defn}

@@

\begin{equation}
\left\{ P(t)\right\} _{ij}=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}\frac{(r t)^{k}}{k!}e^{-r t}=\underset{k=0}{\overset{\infty}{\sum}}\frac{(\mathbf{R}r t)^{k}}{k!}e^{-r t},
\label{eq:matrixExp1}
\end{equation}

\noindent
where $\left\{ P(t)\right\} _{ij}$ denotes the $ij\text{-th}$ element of transition probabilities matrix $\mathbf{P}$ over time $t \geq 0$.
Let us denote the matrix $\mathbf{R}-\mathbf{I}$, where $\mathbf{I}$ is a $K \times K$ identity matrix by $\mathbf{Q}$.
Using the power series definition of the transition probability matrix $\exp(\mathbf{R})=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}/k!$ we can write (\ref{eq:matrixExp1}) as:

\begin{equation}
\left\{ P(t)\right\} _{ij}=e^{\mathbf{R}r t}e^{-r t} =e^{(\mathbf{R}-\mathbf{I})r t}=e^{\mathbf{Q}r t} 
\label{eq:matrixExp2}
\end{equation}

@@

In order to arrive at the finite-time transition probabilities matrix $\mathbf{P}(t)$ for some time $t\geq0$ one needs to compute the matrix exponential in Equation \ref{eq:matrixExp2}.
\cite{Moler78nineteendubious} provide an excellent review of numerical methods to compute the matrix exponent. 
Here we will highlight the method utilized by most phylogenetic software, including BEAGLE \citep{Ayres2012} and BEAST \citep{Drummond2012}, that uses the singular value decomposition (SVD) of the rate matrix $\mathbf{Q}$.

@@

\noindent where $\mathbf{V}$ is a matrix composed of eigenvectors of $\mathbf{Q}$. 
Multiplying both sides of equation (\ref{eq:diagonalization}) with the edge length $t$ and scaling by the appropriate rate scalar $r$, we can write $\mathbf{Q}$ in form:

@@

\noindent 
where the last transition is due to the fact that exponential of diagonal matrix is obtained by exponentiating every entry on the main diagonal.
We should note here that the repeated evaluations of $\exp\left(rt\mathbf{Q}\right)$ for different times $t$ are all based on the same SVD of the matrix $\mathbf{Q}$. 
One needs only to scale and exponentiate the eigenvalues to calculate the diagonal matrix $D$ and perform two matrix multiplications, thus reducing all calculations to a simple and effective algorithm.

@@

The elements of $\mathbf{P}$ quantify the finite-time transition probabilities between the $K$ discrete state-space elements.
We have described here a Continuous-time Markov chain (CTMC), that we will use to describe the substitution process at a single site of an alignment.
Every CTMC is completely characterized by its rate matrix $\mathbf{Q}$ and the stochastic matrix $\mathbf{P}(t),\ t\geq0$ is obtained via matrix exponentiation as described in Subsection \ref{sub:exponentiation}.

@@

Figure \ref{fig:substitution} depicts an example of such a CTMC substitution model, governing the substitution rates between four discrete nucleotide base pairs, where all the pairs communicate (all states have a non-zero transitional probability of exiting from, i.e. all states are non-absorbing).
Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by:

@@

Unfortunately this assumption clearly does not hold for many real-world data-sets, as the temporal dynamics clearly do shift in time, often in a seasonal, observable manner \citep{Bahl2011}.
One of the goals of the research presented in this thesis is the development of a model that relaxes this strict assumption.
% can grasp such situations.

@@

As $t$ goes to infinity the probability that the site is in some state $j$ is non-zero and does not depend on the starting state $i$.
This implies that the chain will always display the same long-term behavior, which is why $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ is called a \emph{stationary distribution} (also referred to as the \emph{equilibrium distribution} or \emph{equilibrium frequencies}).
We can look at the elements of $\mathbf{\Pi}$ as the proportion of time spent in state $j\in\mathcal{E}$ after the Markov chain has run for infinitely long time, that does not depend upon the initial state $i$ of the process.
In other words if we sample the initial state from the stationary distribution and then let the process run for some sufficiently long time $t$, then the distribution of the final state will equal the stationary distribution.

@@

\noindent
Condition \ref{eq:time_reversibility1} is also referred to as the \emph{detailed balance} equation, and posits that the probability of sampling $i$ from the stationary distribution and then going to $j$ over some finite-time $t$ is the same as sampling $j$ and going to $i$. 
% By the definition of $\mathbf{\Pi}$ and $\mathbf{P}$ from (\ref{eq:time_reversibility1}) we have:
% 
% \begin{equation}
% \pi_{i}q_{ij}=\pi_{j}q_{ji}\mathbf{Q}.
% \label{eq:time_reversibility2}
% \end{equation}
% 
% Definition (\ref{eq:time_reversibility1}) and (\ref{eq:time_reversibility2}) are equivalent.
This condition is assumed for strictly computational reasons, i.e. time-reversibility makes it easier to diagonalize $\mathbf{Q}$, which as we have described in 
Subsection \ref{sub:exponentiation} is a strategy frequently employed in phylogenetic software to compute the exponentials of rate matrices.
Time reversibility also simplifies the computation of the likelihood on a tree, since it is then indifferent to the position of the root, which we will discuss in Subsection \ref{sub:likelihood}.

@@

In this subsection we will discuss the simplest model of nucleotide substitution, i.e. the JC69 model \citep{Jukes1969}.
The JC69 model assumes that the instantaneous rates of change between every two nucleotides are the same.
We also require that the rate matrix $Q$ characterizing the model is a stochastic matrix (each row sums to 0).
The infinitestimal rate matrix of the JC69 model is thus given by:

@@

\noindent
The finite-time transition probabilities matrix $\mathbf{P}(t)$ for time $t>0$ is calculated using matrix exponentiation  (see Subsection \ref{sub:exponentiation}):

@@

\noindent
Second, the rate and time in (\ref{eq:jc69Finite}) occur as a product, which means that they are confounded.
If we were to simulate two sequences, using Algorithm listed in (\ref{alg:simulation}) and starting the random number generators form the same seed, they will look the same for every $r$ and $t$ as long as their product $rt$ is the same.
This is true for most of the substitution models and means that typically any inference performed focuses on inferring the product parameter $\theta=rt$.

@@

To gain a better understanding of the process shaping the data observed at the tips of a phylogenetic tree let us consider the problem of simulating a sequence evolution on a tree according to a model with rate matrix $\mathbf{Q}$.
We start with a first character $i$ sampled at the root of the tree $x_0$ from the stationary distribution $\mathbf{\Pi}$.
Let us assume that two directly neighboring nodes are $x_j$ and $x_k$.
The character sampled at the root waits an Exponentialy distributed amount of time $t\geq0$ (see Figure \ref{fig:poisson}) before moving down the tree to one of the children nodes $x_i$.
Such traversal is said to be following the pre-order, i.e. parental nodes are visited before child nodes.   
The process then randomly decides by a coin flip to which of the two neighbouring nodes to move next.
Let us assume that the process moved to $x_j$.
The state $j$ for the node $x_j$ is sampled conditional on the value $i$ already sampled at the root using the branch length expressed in real time $t$ and the vector probabilities in the row $i$ of the finite-time probabilities matrix $\mathbf{P}(t)$, calculated using SVD of the rate matrix $\mathbf{Q}$ that characterizes the CTMC.

@@

This constitutes the Markov property of a CTMC, as defined in (\ref{eq:markov}).
The sampling is then repeated on the node $x_j$, before recursively proceeding to the other trio of nodes.
The states sampled at the tips of the tree constitute the character sequence for one site of the alignment.
In this light the CTMC process can be viewed as a continuous-time random-walk, unfolding on the topology $\mathbf{F}$, with it's behaviour described by a rate-matrix $\mathbf{Q}$.
The complete algorithm is listed in Algorithm \ref{alg:simulation}.

@@

Likelihood is defined as a conditional probability of observing the data given the parameters, in a function of those parameters.
Discrete molecular sequence data comes in a form of aligned matrix $\mathbf{X}$ of size $n \times l$, where $x_{jh}$ is the $h$-th character in $j$-th sequence.
Single column $\mathbf{x}_{h}$ of data matrix constitutes a single site.
Like before the standard assumption of independence posits that different sites evolve independently and that once two branches split at a node, the evolution on those branches is also independent.

In Subsection \ref{sub:evolutionOnTree} we described how evolution according to some process $\mathbf{Q}$ gives rise to the observed data $\mathbf{X}$.
In the typicall situation we start with just the observed data and aim the inference efforts at modelling the process that generated the alignment.
We can therefore denote the likelihood simply as:

@@

In this thesis we will not discuss the maximum likelihood estimation itself, rather focus on closely related methods of Bayesian inference.
The likelihood is nontheless required for both frameworks, and it enters the Bayesian schema as the data component.
We will now discuss how to calculate the likelihood of observing a certain realised sequence of characters on the tips of a tree, given a substitution model with fixed parameters.

@@

To calculate the likelihood means thus to reverse-engineer the evolution as it happened.
Figure \ref{fig:LIKELIHOOD} presents single site with data $TCACC$ observed at the tips of the tree.
External nodes are numbered $x_{1}, \ldots x_{5}$, internal nodes $x_{6}, \ldots x_{8}$ and the root is placed at the node $x_{0}$.
Length of each branches leading to node $i$ is denoted as $t_{i}$.
Using methods presented in Subsection \ref{sub:exponentiation} we calculate the finite-time transition probabilities $p_{ij}(t_{i})$ between all pairs of characters in the alignment.
According to the independence assumptions to calculate the probability at a single site we need to sum over all possible combinations that could give rise to the observed sequence. Thus:

\begin{equation}
\underset{x_{0}\in \mathcal{E}}{\sum}\;\underset{x_{6}\in \mathcal{E}}{\sum}\;\underset{x_{7}\in \mathcal{E}}{\sum}\;\underset{x_{8}\in \mathcal{E}}{\sum}[\pi_{x_{0}}\cdot p_{x_{0}x_{8}}(t_{7})\cdot p_{x_{0}x_{7}}(t_{8})\cdot p_{x_{8}x_{6}}(t_{6})\cdot p_{x_{8}A}(t_{3})\cdot p_{x_{7}C}(t_{4})\cdot p_{x_{7}C}(t_{5})\cdot p_{x_{6}T}(t_{1})\cdot p_{x_{6}C}(t_{2})]
\label{eq:likelihoodNaive}
\end{equation}

We can break down the terms in (\ref{eq:likelihoodNaive}) to the probability of observing the data $TCACC$ for the tips and $x_{0},x_{6},x_{7},x_{8}$ for the ancestral nodes (inside the brackets) and the sum over all possible path combinations over the state space $\mathcal{E}$ (outside the brackets).
Term $\pi_{0}$ is simply the probability that the root has character $x_{0}$ and comes from the assumption that the process generating the data has reached its equilibrium.
